Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-10-06 paper/intro.tex
AEC updates

Introductory programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 1960s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$--$\bReg$ range over thread-local \emph{registers}
and $\aExp$--$\bExp$ range over side-effect-free \emph{expressions}). 
%\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
  \item \label{wp-let}
    $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    %   \stepcounter{enumi}
    % \item[] \labeltext[\textsc{d}2]{}{wp-assign}
    %   \begin{enumerate}[leftmargin=0pt]
    %   \item \label{wp-write}
    %     $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    %   \item \label{wp-let}
    %     $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    %   \item \label{wp-read}
    %     $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
    %   \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
%\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.
%% We have split
%% \citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
%% \eqref{wp-assign} into three cases. In our notation,
%% $\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
%% most once,
%% $\aLoc$--$\cLoc$ range over shared memory references, and
%% $\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
%% include
%% $\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
%%   equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
%%
This is an elegant explanation of sequential computation in a sequential
context. Note that \ref{wp-let} is sound because a read from a thread-local
register must be fulfilled by a preceding write in the same thread.
In a concurrent context, with shared variables
($\aLoc$--$\cLoc$), the obvious generalization
%\begin{multicols}{2}
\begin{itemize}
\item[{\labeltext[\textsc{d}2a]{(\textsc{d}2a)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
\item[{\labeltext[\textsc{d}2b]{(\textsc{d}2b)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
\end{itemize}
%\end{multicols}
\noindent
is unsound!  In particular, a read from a shared memory location may be fulfilled by a write
in another thread, invalidating \ref{wp-read}.  (We assume that expressions
% $\aExp$--$\bExp$
do \emph{not} include shared variables.) % $\aLoc$--$\cLoc$.)

@@

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes (pink arrows \tikz[baseline]{\draw [po] (0,0.1) -- (0.5,0.1);}),
where each read has a matching \emph{fulfilling} write (green arrows \tikz[baseline]{\draw [rf] (0,0.1) -- (0.5,0.1);}),
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \po{wx0}{wx1}
      \po{wx1}{wy2}
      \po{wy2}{ry2}
      \po{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}

@@

This model naturally extends to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \po{wx0}{wx1}
      \po{wx1}{wy2}
      \po{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}

@@

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole program. This allows executions that
are inconsistent with program order, such as:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx0}{\DR{x}{0}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx0}{rx0}
      \po{wx0}{wx1}
      \wk[out=-165,in=-15]{rx0}{wx1}
    \end{tikzinline}}
\end{gather*}

In such models, the causal order between events is important,
and includes control and data dependencies, to avoid
paradoxical ``out of thin air'' examples such as:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}

@@

Other work in relaxed memory has shown that tooling is especially useful to researchers, architects, and language specifiers, enabling them to build intuitions experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
Unfortunately, it is not obvious that tools can be built for all thin-air free models, the calculation of \citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination proof for an arbitrary input, and the enormous state space for the operational models of \citet{DBLP:conf/popl/KangHLVD17} and \citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool builder -- and as yet no tool exists for automatically evaluating these models.
We describe a tool, \PwTer, for automatically evaluating \PwT{} in \textsection\ref{sec:tool}.

>>>>

Novice programmers are taught \emph{sequential abstraction}: that the
program $\aCmd_1\SEMI \aCmd_2$ executes $\aCmd_1$ before $\aCmd_2$.  Since
the late 1960s, we've been able to explain this using logic
\citep{Hoare:1969:ABC:363235.363259}.  In
\citeauthor{DBLP:journals/cacm/Dijkstra75}'s
[\citeyear{DBLP:journals/cacm/Dijkstra75}] formulation, we think of programs
as \emph{predicate transformers}, where predicates describe the state of
memory in the system.  In the calculus of weakest preconditions, programs map
postconditions to preconditions.  We recall the definition of
$\fwp{\aCmd}{\bForm}$ for loop-free code below (where $\aReg$--$\bReg$ range over thread-local \emph{registers}
and $\aExp$--$\bExp$ range over side-effect-free \emph{expressions}). 
%\begin{multicols}{2}
  \begin{enumerate}[,label=(\textsc{d}\arabic*),ref=\textsc{d}\arabic*]
  \item
    \begin{math}
      \fwp{\SKIP}{\bForm} = \bForm
    \end{math}
  \item \label{wp-let}
    $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    % \item 
    %   \begin{math}
    %     \fwp{\ABORT}{\bForm} = \FALSE
    %   \end{math}
    %   \stepcounter{enumi}
    % \item[] \labeltext[\textsc{d}2]{}{wp-assign}
    %   \begin{enumerate}[leftmargin=0pt]
    %   \item \label{wp-write}
    %     $\fwp{\PW{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
    %   \item \label{wp-let}
    %     $\fwp{\LET{\aReg}{\aExp}}{\bForm} = \bForm[\aExp/\aReg]$
    %   \item \label{wp-read}
    %     $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \aLoc{=}\aReg\limplies\bForm$ %\bForm[\aLoc/\aReg]$ %
    %   \end{enumerate}
    % \item 
    %   \begin{math}
    %     \fwp{\LET{x}{\aExp}}{\bForm} = \bForm[\aExp/x]
    %   \end{math}
  \item
    \begin{math}
      \fwp{\aCmd_1;\aCmd_2}{\bForm} = \fwp{\aCmd_1}{\fwp{\aCmd_2}{\bForm}}
    \end{math}
  \item
    \begin{math}
      \fwp{\IF{\aExp}\THEN \aCmd_1\ELSE \aCmd_2\FI}{\bForm}= {}
    \end{math}
    \begin{math}
      ((\aExp{\ne}0) \limplies \fwp{\aCmd_1}{\bForm}) \land
      ((\aExp{=}0) \limplies \fwp{\aCmd_2}{\bForm})
    \end{math}
  \end{enumerate}
%\end{multicols}
For this language, the Hoare triple
$\hoare{\aForm}{\aCmd}{\bForm}$ holds exactly when $\aForm \limplies
\fwp{\aCmd}{\bForm}$.
%% We have split
%% \citeauthor{DBLP:journals/cacm/Dijkstra75}'s rule for assignment
%% \eqref{wp-assign} into three cases. In our notation,
%% $\aReg$--$\bReg$ range over thread-local registers, which may be assigned at
%% most once,
%% $\aLoc$--$\cLoc$ range over shared memory references, and
%% $\aExp$--$\bExp$ range over thread-local expressions, which do \emph{not}
%% include
%% $\aLoc$--$\cLoc$.\footnote{Under these assumptions, \eqref{wp-read} is
%%   equivalent to $\fwp{\PR{\aLoc}{\aReg}}{\bForm} = \bForm[\aLoc/\aReg]$.}
%%
This is an elegant explanation of sequential computation in a sequential
context. Note that \ref{wp-let} is sound because a read from a thread-local
register must be fulfilled by a preceding write in the same thread.
In a concurrent context, with shared variables
($\aLoc$--$\cLoc$), the obvious generalization
%\begin{multicols}{2}
\begin{itemize}
\item[{\labeltext[\textsc{d}2a]{(\textsc{d}2a)}{wp-write}}] $\fwp{\LET{\aLoc}{\aExp}}{\bForm} = \bForm[\aExp/\aLoc]$
\item[{\labeltext[\textsc{d}2b]{(\textsc{d}2b)}{wp-read}}] $\fwp{\LET{\aReg}{\aLoc}}{\bForm} = \bForm[\aLoc/\aReg]$
\end{itemize}
%\end{multicols}
\noindent
is unsound!  In particular, a read from a shared memory location may be fulfilled by a write
in another thread, invalidating \ref{wp-read}.  (We assume that expressions
% $\aExp$--$\bExp$
do \emph{not} include shared variables.) % $\aLoc$--$\cLoc$.)

@@

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order over memory actions ($\xwki$),
where each read has a matching \emph{fulfilling} write ($\xrf$),
for example:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}\SEMI
    \PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{wy2}{ry2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}

@@

This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
In general, there will be many such executions, reflecting different
interleavings of the threads.

@@

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location}. This allows executions that
are inconsistent with program order, such as:
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx0}{\DR{x}{0}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx0}{rx0}
      \wki{wx0}{wx1}
      \wk[out=-165,in=-15]{rx0}{wx1}
    \end{tikzinline}}
\end{gather*}
Reads must be ordered between the available writes, creating \emph{antidependencies} $(\DR{x}{0}\xwk\DW{x}{1})$.

In such models, the causal order between events is important,
and includes control and data dependencies ($\xpo$), to avoid
paradoxical ``out of thin air'' examples such as:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \IF r \THEN \PW{y}{1} \FI}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{rx1}{wy1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
\begin{gather*}
  \THREAD{\PR{x}{r}\SEMI \PW{y}{1}}
  \PAR
  \THREAD{\PR{y}{s}\SEMI \PW{x}{s}}
  \\[-.4ex]
  \nonumber
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{rx1}{\DR{x}{1}}{}
      \event{wy1}{\DW{y}{1}}{right=of rx1}
      \event{ry1}{\DR{y}{1}}{right=3em of wy1}
      \event{wx1}{\DW{x}{1}}{right=of ry1}
      \rf[out=15,in=165]{wy1}{ry1}
      \rf[out=165,in=15]{wx1}{rx1}
      \po{ry1}{wx1}
    \end{tikzinline}}
\end{gather*}

@@

Other work in relaxed memory has shown that tooling is especially useful to researchers, architects, and language specifiers, enabling them to build intuitions experimentally~\cite{DBLP:conf/esop/PaviottiCPWOB20,PrideMM,alglave,Batty:2011:MCC:1926385.1926394}.
Unfortunately, it is not obvious that tools can be built for all thin-air-free models, the calculation of \citet{Pichon-Pharabod:2016:CSR:2837614.2837616} does not have a termination proof for an arbitrary input, and the enormous state space for the operational models of \citet{DBLP:conf/popl/KangHLVD17} and \citet{DBLP:journals/pacmpl/ChakrabortyV19} is a daunting prospect for a tool builder -- and as yet no tool exists for automatically evaluating these models.
We describe a tool, \PwTer, for automatically evaluating \PwT{} in \textsection\ref{sec:tool}.
